|

### 2019-08-10 article/article.Rmd
update

One of the key determinants of efficient public services provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability has a positive impact on governance, as it ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Recent research also suggests accountability leads to higher economic growth because it limits state discretion in the economy and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

However, accountability systems take many forms. One promising model is that of _bottom-up monitoring_, in which citizens receive information about the shortcomings of a given project so they can evaluate and pressure underperforming public officials [@kosack2014does; @molina2016community; @raffler2018weakness]. Proponents argue bottom-up accountability is effective because: 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to attack corruption that directly affects them; 3) policy-makers are sensitive to social punishment from their own communities [@serra2011combining, 570]. In a nutshell, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of state officials with those of the constituency they serve [@barro1973control; @raffler2018weakness, 2].

@@

[^transparencia]: Transparência Brasil is an non-governmental organisation whose mission is to 'promote transparency and social control of public power' in Brazil. It has been active since April 2000, is non-partisan and receives no public funding. More information at <http://transparencia.org.br> (access: July 2019).

@@

The _Tá de Pé_[^tdp] cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve responsiveness in government education expenditures. The TDP app incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and that information will be assessed by a group of independent specialists. In the case the construction is behind schedule, TDP provides a writing platform whereby citizens can contact their representatives quickly and anonymously. The app then prompts citizen to write a notification to the mayor's office, which has 15 days to reply. If they do not respond to the request, the app forwards the notification to the Brazilian federal government, making it harder to the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsive to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS appeared six months later. A massive Facebook campaign started in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns were concentrated in October 2017 and cost US$ 4,479.29. Facebook advertisements attracted 2,028 new users to the platform in October 2017 only.

@@

Transparência Brasil partnered with the Brazilian branch of _Engineers without Borders_ (EWB), an independent non-governmental organisation[^ewb], to provide technical assessment of school completion rates. The EWB bases their evaluations on photos and GPS location sent through the app. The engineers' reports are later uploaded to the database and stored on the users' mobile devices so citizens can track the progress of the constructions.

@@

The TDP has also received feedback from Brazilian computer programmers and policy analists. In 2017 and 2018, Transparência Brasil announced two team programming competitions, called 'Tá de Pé Hackathons', where contributors could fix code bugs and suggest new functionalities to the TDP project. One of these innovations consists of a Twitter bot (<https://twitter.com/tadepeapp>) which posts a message on the social network each time a user submits a new picture for evaluation. This allows interested parties to track the progress of construction sites as well as add new public works to the TDP database.

@@

Between August 2017 and August 2018, we implemented two interventions to measure the effect of the TDP app on six school construction outcomes. The outcomes are: 1) the percentage of public investment allocated to school constructions; 2) the difference between the percentage invested before and after the interventions; 3) the number of finished constructions; 4) the number of cancelled constructions; 5) the number of updated conclusion dates; 6) a placebo outcome indicating the percentage of the investment executed before the impact evaluation started. We expect the first five factors to increase the last one to remain unchanged.

The first intervention was carried out from August 2017 to February 2018 using the Android version of the TDP app. The unit of analysis for this experiment is the Brazilian municipality. We randomly selected 150 cities to the control group and included 1030 municipalities in the treatment group. Our control condition consisted in removing all school constructions in the chosen municipalities so that citizens were unable to add information about them.

@@

We also conducted two manipulation checks and analysed the number of TDP app downloads by municipality and app downloads over time. Figure \ref{fig:manipulation1} displays the results. The checks display good territorial variability. There were 455 downloads in the 1030 municipalities in the treatment condition. Downloads peak during the Facebook TDP campaign, right after the TDP app launch in August, then diminish over time. Please refer to the Supplementary Materials for a detailed description of the tests.

```{r man1, echo=FALSE, warning=FALSE, messages=FALSE, cache=TRUE, error=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center', fig.cap='\\label{fig:manipulation1}Manipulation checks for intervention 1. The first plot shows the geographical distribution of the treatment condition and the second graph displays the number of TDP app downloads from August 2017 to February 2018.'}
# Load data 
load('tdp.RData')

@@

The second intervention is similar to intervention 1 in all but three characteristics. First, the TDP app was available for Android and iOS devices. Also, we randomised the intervention at the school level, with 654 control and 3895 treatment units. We used blocked randomisation stratified by Brazilian state, school construction status (under construction, stopped, unfinished) and whether the municipality spent more on school construction that the distribution median. Finally, the intervention period lasted from August 2018 to February 2019.

@@

$$ Y_i \ = \ \alpha + \beta T_i + \gamma X_i + \theta Z_i + \varepsilon_i $$

where $i$ indexes the experiment units. $Y_i$ in one of the six outcomes described above and $\beta$ denotes the average treatment effect. $T_i$ is a binary treatment indicator, $\gamma$ is a vector of fixed effects, $X_i$ is a matrix of Brazilian states' fixed effects, $\theta$ is a vector of controls and $Z_i$ an array of controls for the case $i$. The error term is denoted by $\varepsilon_i$. We cluster the standard errors at the municipal level as mayors are responsible for school investment decisions.

@@

** TABLE 1 ABOUT HERE -- MAIN RESULTS INTERVENTION 1 **

>>>>

One of the key determinants of efficient public services provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability has a positive impact on governance, as it ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Recent research also suggests that accountability leads to higher economic growth because it limits state discretion in the economy and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

However, accountability systems take many forms. One promising model is that of _bottom-up monitoring_, in which citizens receive information about the shortcomings of a given project so they can evaluate and pressure underperforming public officials [@kosack2014does; @molina2016community; @raffler2018weakness]. Proponents argue bottom-up accountability is effective because: 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to attack corruption that directly affects them; 3) policy-makers are sensitive to social punishment from their own communities [@serra2011combining, 570]. In this regard, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of state officials with those of the constituency they serve [@barro1973control; @raffler2018weakness, 2].

@@

[^transparencia]: Transparência Brasil is an non-governmental organisation whose mission is to 'promote transparency and social control of public power'. It has been active since April 2000, is non-partisan and receives no public funding. More information at <http://transparencia.org.br> (access: July 2019).

@@

The _Tá de Pé_[^tdp] (TDP) cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve responsiveness in government education expenditures. The TDP app incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and that information will be assessed by a group of independent specialists. In the case the construction is behind schedule, TDP provides a writing platform whereby citizens can contact their representatives quickly and anonymously. The app then prompts citizens to write a notification to the mayor's office, which has 15 days to reply. If they do not respond to the request, the app forwards the notification to the Brazilian federal government, making it harder to the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsive to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS appeared six months later. In October 2017, we started a Facebook campaign in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns cost US$ 4,479.29 and Facebook advertisements attracted 2,028 new users to the platform in October 2017 only.

@@

Transparência Brasil partnered with the Brazilian branch of _Engineers without Borders_ (EWB), an independent non-governmental organisation[^ewb], to provide technical assessment of school completion rates. The EWB bases their evaluations on photos and GPS coordinates sent through the app. The engineers' reports are later uploaded to the TDP database and stored on the users' mobile devices so citizens can follow the progress of the constructions.

@@

The TDP has also received feedback from Brazilian computer programmers and policy analists. In 2017 and 2018, Transparência Brasil announced two team programming competitions, called 'Tá de Pé Hackathons', where contributors could fix code bugs and suggest new functionalities to the TDP project. One of these innovations consists of a Twitter bot (<https://twitter.com/tadepeapp>) which posts a message on the social network each time a user submits a new picture for evaluation. This allows any interested parties, including those who do not use the TDP app, to check the state of school construction sites.

@@

Between August 2017 and August 2018, we implemented two interventions to measure the effect of the TDP app on six school construction outcomes. The outcomes are: 1) the percentage of public investment allocated to school constructions; 2) the difference between the percentage invested before and after the interventions; 3) the number of finished constructions; 4) the number of cancelled constructions; 5) the number of updated conclusion dates; 6) a placebo outcome indicating the percentage of the investment executed before the impact evaluation started. We expect the first five factors to increase and the last one to remain unchanged.

The first intervention was carried out from August 2017 to February 2018 using the Android version of TDP. The unit of analysis for this experiment is the Brazilian municipality. We randomly selected 150 cities to the control group and included 1030 municipalities in the treatment group. Our control condition consisted in removing all school constructions in the chosen municipalities so that citizens were unable to add information about them.

@@

We also conducted two manipulation checks and analysed the number of TDP app downloads by municipality and over time. Figure \ref{fig:manipulation1} displays the results and indicates that the treatment has good territorial variability. There are 455 downloads in the 1030 municipalities in the treatment condition. Downloads peak during the Facebook TDP campaign, launched in October 2017, then diminish in the following months.

```{r man1, echo=FALSE, warning=FALSE, messages=FALSE, cache=TRUE, error=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center', fig.cap='\\label{fig:manipulation1}Manipulation checks for intervention 1. The first plot shows the geographical distribution of the treatment condition and the second graph displays the number of TDP app downloads from August 2017 to February 2018.'}

# Load data 
load('tdp.RData')

@@

The second intervention is similar to intervention 1 in all but three characteristics. First, the TDP app was available for both Android and iOS devices. Second, we randomised the intervention at the school level, with 654 control and 3895 treatment units. We used blocked randomisation stratified by Brazilian states, school construction status (under construction, stopped, unfinished) and whether the municipality spent more on school construction that the distribution median. Finally, the intervention period lasted from August 2018 to February 2019.

@@

\begin{equation} 
Y_i \ = \ \alpha + \beta T_i + \gamma X_i + \theta Z_i + \varepsilon_i 
\end{equation}

where $i$ indexes the experiment units. $Y_i$ in one of the six outcomes described above, $\alpha$ is the intercept, $\beta$ denotes the average treatment effect and $T_i$ is a binary treatment indicator. $\gamma$ is a vector of fixed effects, $X_i$ is a matrix of Brazilian states' fixed effects, $\theta$ is a vector of controls and $Z_i$ an array of controls for the case $i$. The error term is denoted by $\varepsilon_i$. We cluster the standard errors at the municipality level as mayors are responsible for school investment decisions in Brazil.

@@
